Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-05-30 submission/response_to_reviewers.md

We appreciate the comments provided by the two reviewers and are grateful for their suggestions and overall positive view of the manuscript. We have done our best to address their comments below. Where appropriate, we have referenced line numbers in the marked up version of the manuscript.

**Reviewer #1 (Comments for the Author):**

**This paper investigates associations between colon cancer and fecal microbiota/SCFAs. This is an important study because intersections between each of these variables are current active areas of research in the field. Overall, I found this paper concise and easy to read. I also found the analyses clear. My primary comments (described below) center on the interpretation of those analyses.**

**Major points:**

**This manuscript deals with the tricky case of how to report non-significant (i.e. p>0.05) results. The approach the authors take is to argue for a lack of association between the measured variables (e.g. fecal SCFAs & tumor status). I have two concerns about this interpretation. First, statisticians are increasingly clarifying that the lack of a significant statistical result does not mean there is no association between two variables (see Amrhein et al., Nature 2019). Second, p-values reflect the odds that a given statistical result was arrived at by chance, assuming the null hypothesis is true. Those odds, in my opinion, are not fundamentally different between 0.05 and say 0.058 or even 0.077 (L66; L70). Additionally, the authors even report an instance where a model associating SCFAs and tumor status has a p-value below the conventional cutoff (p < 0.0001; L79,80).**

**Still, I don't mean to encourage alternative problematic interpretations such as trumpeting significant associations when p-values are marginal. I think instead what might be merited here is nuance. Perhaps rather than rejecting an association, the authors could consider simply reporting the p-values, as well as a measure of effect (e.g. difference of means or odds ratios). Those effect sizes, I suspect, will support the conclusion that even if associations exist, they are weak (which is how the random forest models in L72-85 are presented).**

> We agree that relying heavily on an arbitrary threshold for establishing significance or lack of significance is problematic. This was why we also showed the data in Figure 1. These box plots hopefully make it clear that although there may be a couple of nearly significant P-values, the effect sizes are small and are inconsistent. We have added a sentence to emphasize this point at L75-77. We have considered different ways of reporting the P-values and effect sizes and have concluded that the revised approach is the most suitable given the word length restrictions on an Observation format manuscript.

> For the reviewer's benefit in the paragraph from L63 to L78...

> * The "all P>0.148" set of P-values for the cross sectional data included values of ***0.148***, 0.222, and 0.655 for acetate, butyrate, and propionate, respectively. The largest effect sizes for these three comparisons were 2.45, 1.72, and 0.352 mmol/kg
> * The "all P>0.058" set of P-values for the follow-up data included values of 0.508 [cancer] and 0.350 [adenoma], 0.916 [cancer] and 0.101 [adenoma], 0.853 [cancer] and ***0.058*** [adenoma] for acetate, butyrate, and propionate, respectively. The corresponding differences between the before and after samples were 3.00 [cancer] and -3.84 [adenoma], 0.312 [cancer] and -1.75 [adenoma], -0.396 [cancer] and -1.68 [adenoma] for acetate, butyrate, and propionate, respectively.
> * The "all P>0.16" set of P-values for the relative concentrations of SCFA included values of 0.571, 0.765, and 0.245 for acetate, butyrate, and propionate, respectively from the cross sectional data. They were ***0.164*** [cancer] and 0.322 [adenoma], 0.692 [cancer] and 0.432 [adenoma], and 0.210 [cancer] and 0.365 [adenoma] for acetate, butyrate, and propionate follow-up data, respectively. The largest effect sizes for the cross-sectional data were 0.006, 0.004, and 0.011  for acetate, butyrate, and propionate, respectively. The differences in relative SCFA concentrations in the follow-up data were 0.029 [cancer] and 0.015 [adenoma], -0.006 [cancer] and -0.007 [adenoma], and -0.020 [cancer] and -0.011 [adenoma] for acetate, butyrate, and propionate, respectively.
> * The "all P>0.077" set of P-values for the total concentration of SCFA included values of 0.189 for the cross sectional data and 0.833 for the follow-up cancer data and 0.150 for the follow-up adenoma data. In addition the set included values for the per-molecule of carbon basis which were 0.216 for the cross sectional data and 0.916 for the follow-up cancer data and ***0.077*** for the follow-up adenoma data.

> Again, aside from cluttering the figure with P-values or the text with extra effect sizes we think it is best to leave it to the reader to see that the effect sizes in Figure 1 are quite small.

> Finally, the reviewer points out that "a model associating SCFAs and tumor status has a p-value below the conventional cutoff (p < 0.0001; L79,80)". We point out at that location in the manuscript that the comparison is to a random classification and that the actual AUROC was only 0.54 and 0.55, which was unimpressive. We feel that this was sufficient to explain the P-value.

**I encourage the authors to consider tempering their language regarding the association between fiber intake, SCFAs and protection tumor risk (e.g. L18). I am aware that butyrate and other SCFAs may be over-hyped as a panacea for GI disease. Still, if protection is mediated by SCFA uptake (to me the most reasonable model) then excreted (i.e. fecal SCFAs) should not necessarily be correlated with disease. Additionally, the authors are examining a single point measurement of SCFAs, without consideration of how noisy they may be. It's possible that more frequent longitudinal measurements would reveal a stronger association.**

> We appreciate the reviewer's concern that we are over interpreting negative results. In various places where we provide background to the problem, including L18, we are pointing to epidemiological data which provide a confused picture of the value of fiber and SCFAs in preventing colorectal cancer (CRC). We have gone back through the reporting of our results to emphasize the use of fecal SCFAs, a single time point, and the goal of classification. The most significant edits were made at L31-32, L114, and L118. For example, at L31-33, we rewrote, *"These data indicate that there is no conclusive link between the gut microbiome, SCFAs, and tumor burden."* as *"Our results indicate that the association between fecal SCFAs, the gut microbiome, and tumor burden is weak."*

> It should be noted that our manuscript is in the context of diagnosis, not necessarily identifying mechanisms of tumorigenesis that are mediated by microorganisms. We are limited in the amount of discussion we can provide because of word limitations, but it is important to note that it is technically challenging/impossible to obtain mucosal SCFA concentrations since the typical bowel preparation strips away much of the mucosa limiting the ability to sample the SCFAs at the interface. Furthermore, considering our goal is to develop a diagnostic to detect colonic lesions, longitudinal sampling is unlikely to be possible.

**Minor points: **

**L38: This line could be interpreted as saying any bacterium could cause mutations, inflammations, etc. But, to date, only select bacteria are known to cause such events.**

> We have revised the text from "...role as bacteria are known to..." to "...role as some bacteria are known to..."

**L45: "SCFA have ... anti-proliferative activities." This is true but only contextually. Butyrate in particular has proliferative activity in colonocytes. (See Donohoe et al. Mol Cell. The Warburg Effect Dictates the Mehcanism of Butyrate Mediated Histone Acetylation and Cell Proliferation. 2012)**

> This sentence has been edited to highlight the contextual nature of this observation.

**L55: Insuring should be 'ensuring'**

> This as been corrected as suggested.

**L65: Is it really surprising that there are no effects 1 year after treatment? Also, what exactly were the treatments?**

> We have added text to the sentence to make this more clear at L70 - "None of the *individuals showed signs of recurrence and yet none of the* SCFAs exhibited a significant change with treatment (all P>0.058; Figure 1B)". We previously showed that the microbiome of those individuals diagnosed with carcinomas returned to a more normal state following treatment. If SCFAs were associated with disease, then the prediction would be that the concentrations would change with a return to health. That we did not see such a change further supports our assertion that there is a tenuous link between fecal SCFA concentrations and health status.

> Because of the limits to the number of words available in the *mBio* Observation Format, we do not have the space to recap the treatments. Patients underwent surgical resection and/or chemotherapy. This is more extensively covered in the referenced paper by Sze et al. (2017) [citation 17].

**L92: To be consistent with other statistical tests reported in this manuscript, can the authors provide some measure of how often an R2 of 0.14 would be achieved by chance?**

> The R2 values that we report are the median observed values for our 80/20 split hold out permutation test based on 100 permutations. The distribution of observed values are provided in Supplemental Figure 2. If the association were truly random, the R2 value would be near zero. Our point is that the observed R2 values are quite low (e.g. at most 0.14) and although they are significantly different from zero, the models do a poor job of predicting SCFA concentrations. Because we are not comparing models to each other (e.g. an OTU-based model to an OPF-based model), there is no P-value to compare across R2 values.

**L135: The authors use an aqueous extraction and aqueous injection method for measurement of SCFA by HPLC, and do not acidify the aqueous solution to a pH of ~3 before filtration, as has previously been described as essential for accurate quantitation by GC (see Zhao, Nyman, Jönsson. Biomedical Chromatorgraphy (2006)), and remains a standard practice in GC (see Teixeira et al, British Journal of Nutrition (2013)). Why would acidification of the sample to ensure protonation of the SCFA not be necessary for efficient extraction, as the anionic form of the conjugate base is extracted less efficiently from the stool matrix? The literature often does not provide a rationale for the lack of a need for acidification for quantitation of SCFA from feces. If such a rationale has been published, can the authors please provide this reference?**

> We are unaware of a specific reference as to why SCFAs are extracted differently when using GC or HPLC. The protocol we used is based on that used by Venkataraman et al. (2016)[citation 19], which developed their methods based on the existing literature. The approach is effective and has worked well for a variety of fecal samples.

**L173: Can the authors define 'mtry'? Some readers might not be familiar with this term.**

> We have added a sentence at L183-185 to clarify the meaning of this parameter, *"The mtry parameter represents the number of features randomly sampled from the available features at a question point in the classification tree (i.e. called splits of nodes) that, when answered, lead to the greatest improvement in classification"*.

**Fig. 1B: Are the units here the same as in Fig 1A? Can these be specified?**

> They are the same units and we've inserted them with the y-axis label for Figure 1B.

**Reviewer #2 (Comments for the Author):**

**This manuscript by Sze et al., carries out testing on fecal SCFAs, fecal 16S/metagenomics, in order to assess the value of SCFAs to a colorectal cancer classifier. In this aspect, it is a fine contribution and of interest to the readers of mBio. On that basis I like it, with one major caveat:**

**The language of the claims is problematic. It is very hard to prove that something cannot be done. I would have less of a problem with these statements if there were multiple methods or an attempt to quantify the overall amount of information embedded in SCFAs to try to find an upper bound for how predictive they could be. Instead, the authors try one or two tests in each case and declare that SCFAs cannot be used for CRC classification. In other words, the scientific diligence is not commensurate with the broad nature of the claims.**

> This reviewer shares the concerns of the first reviewer, which we have hopefully addressed. We agree that it is impossible to prove that something cannot be done - scientifically we can't *prove* anything. We can only propose hypotheses and either reject them or find support for the hypothesis.

> We disagree with the reviewer's comment that we haven't pursued multiple methods. First of all, we used a large collection of samples from a cross sectional study and a collection of samples from people with a significant colonic lesion who underwent treatment. Second, our analysis compared SCFA concentrations across diagnosis groups with straightforward statistical tests and we used those data to see if they would improve the ability to diagnose individuals using 16S rRNA gene sequence data, metagenomic data, and imputed metagenomic sequence data. Finally, we attempted to use the 16S rRNA gene sequence data, metagenomic data, and imputed metagenomic sequence data to predict the fecal SCFA concentrations. For these comparisons we tested for an association using the individual concentrations of each SCFA, the relative concentration of each SCFA, the total concentration of SCFAs, and the total amount of SCFA on a per carbon molecule basis. Although, the only classification modeling approach we used was random forest; however, this method and our testing framework are considered to be less sensitive to overfitting, able to incorporate non-linearities in the data, and treat the data in a context-dependent manner. Other methods are available (e.g. logistic regression, SVM, decision tree, etc.), but at least in our hands with similar data do not perform as well as random forest.

**More specifically, I would suggest the following edits to reign this in:**

**1. "Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas" should be changed to "Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas in a random forest machine learning model."**

**2. "Knowledge of microbial community structure does not predict SCFA concentrations in a random forest regression model."**

> Both of these section headings have been edited as suggested by the reviewer.

### 2019-05-30 submission/manuscript.Rmd
Fix email address
Make figures an explicit dependency and remove from body of manuscript
Respond to reviewers' comments

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

@@

Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation. Short-chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber. We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data. We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no significant associations between SCFA concentration and tumor status. When we incorporated these concentrations into random forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals. Finally, we generated random forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples. These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations. These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer. Although other bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited predictive power.

@@

Considering colorectal cancer is the third leading cancer-related cause of death within the United States, there is a great need to detect colorectal tumors early without invasive colonoscopy procedures and to prevent the formation of tumors. Short-chain fatty acids (SCFAs) are often used as a surrogate for measuring gut health and for being anti-carcinogenic because of their anti-inflammatory properties. We evaluated the fecal SCFA concentration of a cohort of individuals with varying colonic tumor burden who were previously analyzed to identify microbiome-based biomarkers of tumors. We were unable to find an association between SCFA concentration and tumor burden or use SCFAs to improve our microbiome-based models of classifying people based on their tumor status. Furthermore, we were unable to find an association between the fecal community structure and SCFA concentrations. These data indicate that there is no conclusive link between the gut microbiome, SCFAs, and tumor burden.

@@

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental, behavioral, and dietary factors [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to have a role as bacteria are known to cause mutations, induce inflammation, and accelerate tumorigenesis [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

Many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016]. SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016]. Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018]. These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988]. In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018]. Such studies are plagued by difficulties insuring subjects took the proper dose and using subjects with prior polyp history who may be beyond a point of benefiting from fiber supplementation. Together, these findings temper enthusiasm for treatments that target the production of SCFAs or for using them as biomarkers for protection against tumorigenesis.

@@

**Fecal SCFA concentrations did not vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016]. We were unable to detect a significant difference in any SCFA concentration across the diagnoses groups (all P>`r p_cross_min`; Figure 1A). Among the individuals with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017]. None of the SCFAs exhibited a significant change with treatment (all P>`r p_follow_min`; Figure 1B). For both the pre-treatment cross-sectional data and the pre/post treatment data, we also failed to detect any significant differences in the relative concentrations of any SCFAs (P>`r rel_test`). Finally, we pooled the SCFA concentrations on a total and per molecule of carbon basis and again failed to observe any significant differences (P>`r pooled_test`). These results demonstrated that there were no significant associations between fecal SCFA concentration and diagnosis or treatment.

@@

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using random forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1). Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001). However, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively, indicating that SCFAs had poor predictive power on their own (Figure 2A). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the same models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a subject's fecal sample did not improve the ability to diagnose a colonic lesion.

@@

**Knowledge of microbial community structure does not predict SCFA concentrations.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis. We trained random forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2). The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared`, which was observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B). We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained random forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2). Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration. The largest R^2^ was `r top_metagenomics$median_Rsquared`, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2). SCFA concentrations could not be predicted based on the imputed metagenomic data. The largest R^2^ was `r top_picrust$median_Rsquared`, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.

**Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions. Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive of SCFA concentrations. These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors. It is important to note that our analysis was based on characterizations of SCFA and microbiome profiles using fecal samples and that observations along the mucosa near the site of lesions may provide a stronger association. This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally. Going forward it is critical to develop additional hypotheses for how the microbiome and host interact to drive tumorigenesis so that we can better understand tumorigenesis and identify biomarkers that will allow early detection of lesions.

@@

**random forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations. The regression models were built to predict the SCFA concentrations of acetate, butyrate, and propionate from microbiome information. For classification and regression models, we pre-processed the features by scaling them to vary between zero and one. Features with no variance in the training set were removed from both the training and testing sets. We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data. The training set was used for hyperparameter selection and training the model and the test set was used for evaluating prediction performance. For each model, the best performing hyperparameter, mtry, was selected in an internal five-fold cross-validation of the training set with 100 randomizations. Six values of mtry were tested and the value that provided the largest AUROC or R^2^ was selected. We trained the random forest model using the selected mtry value and predicted the held-out test set. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance. We used AUROC and R^2^ as the prediction performance metric for classification and regression models, respectively. We used randomForest package implemented to the caret package (version `r installed.packages()["randomForest" ,"Version"]`) in R statistical software (version `r installed.packages()["caret" ,"Version"]`) for our models.

@@

**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Sze_SCFACRC_XXXX_2019/.

@@

\includegraphics{figure_1.png}

@@

\includegraphics{figure_2.png}

@@

\includegraphics{figure_s1.png}

@@

\includegraphics{figure_s2.png}

>>>>

$\dagger$ To whom correspondence should be addressed: \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

@@

Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation. Short-chain fatty acids (SCFAs) have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber. We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data. We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no significant associations between SCFA concentration and tumor status. When we incorporated these concentrations into random forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals. Finally, we generated random forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples. These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations. These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer. Although other bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited predictive power.

@@

Considering colorectal cancer is the third leading cancer-related cause of death within the United States, it is important to detect colorectal tumors early and to prevent the formation of tumors. Short-chain fatty acids (SCFAs) are often used as a surrogate for measuring gut health and for being anti-carcinogenic because of their anti-inflammatory properties. We evaluated the fecal SCFA concentration of a cohort of individuals with varying colonic tumor burden who were previously analyzed to identify microbiome-based biomarkers of tumors. We were unable to find an association between SCFA concentration and tumor burden or use SCFAs to improve our microbiome-based models of classifying people based on their tumor status. Furthermore, we were unable to find an association between the fecal community structure and SCFA concentrations. Our results indicate that the association between fecal SCFAs, the gut microbiome, and tumor burden is weak.

@@

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental, behavioral, and dietary factors [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to have a role as some bacteria are known to cause mutations, induce inflammation, and accelerate tumorigenesis [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

Many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016]. Depending on the concentrations, SCFAs can have proliferative activities at low concentrations or anti-proliferative activities at higher concentrations; they can also have anti-inflammatory activities [@OKeefe2016]. Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018]. These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs, may confer beneficial effects against colorectal cancer. Regardless, there is a lack of consistent evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988]. In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018]. Such studies are confounded by difficulties ensuring subjects took the proper dose and using subjects with prior polyp history who may be beyond a point of benefiting from fiber supplementation. Together, these findings temper enthusiasm for treatments that target the production of SCFAs or for using them as biomarkers for protection against tumorigenesis.

@@

**Fecal SCFA concentrations did not vary with diagnosis or treatment.** To test for a significant association between colorectal cancer and SCFAs, we quantified the concentration of acetate, propionate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016]. We were unable to detect a significant difference in any SCFA concentration across the diagnoses groups (all P>`r p_cross_min`; Figure 1A). Among the individuals with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017]. None of the individuals showed signs of recurrence and yet none of the SCFAs exhibited a significant change with treatment (all P>`r p_follow_min`; Figure 1B). For both the pre-treatment cross-sectional data and the pre/post treatment data, we also failed to detect any significant differences in the relative concentrations of any SCFAs (P>`r rel_test`). Finally, we pooled the SCFA concentrations on a total and per molecule of carbon basis and again failed to observe any significant differences (P>`r pooled_test`). Although some of the P-values from our analyses were close to 0.05, the effect sizes were all relatively small and inconsistent given the disease progression (Figure 1). These results demonstrated that there were no significant associations between fecal SCFA concentration and diagnosis or treatment.

@@

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas using a random forest model.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using random forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1). Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001). However, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively, indicating that SCFAs had poor predictive power on their own (Figure 2A). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the same models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a subject's fecal sample did not improve the ability to diagnose a colonic lesion.

@@

**Knowledge of microbial community structure does not predict SCFA concentrations using a random forest model.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis. We trained random forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2). The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared`, which was observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B). We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained random forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2). Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration. The largest R^2^ was `r top_metagenomics$median_Rsquared`, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2). SCFA concentrations could not be predicted based on the imputed metagenomic data. The largest R^2^ was `r top_picrust$median_Rsquared`, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict fecal SCFA concentrations.

**Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions. Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not meaningfully predictive of SCFA concentrations. These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors. It is important to note that our analysis was based on characterizations of SCFA and microbiome profiles using fecal samples at a single time point. Furthermore, observations along the mucosa near the site of lesions may provide a stronger association. This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally. Going forward it is critical to develop additional hypotheses for how the microbiome and host interact to drive tumorigenesis so that we can better understand tumorigenesis and identify biomarkers that will allow early detection of lesions.

@@

**Random forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations. The regression models were built to predict the SCFA concentrations of acetate, butyrate, and propionate from microbiome information. For classification and regression models, we pre-processed the features by scaling them to vary between zero and one. Features with no variance in the training set were removed from both the training and testing sets. We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data. The training set was used for hyperparameter selection and training the model and the test set was used for evaluating prediction performance. For each model, the best performing hyperparameter, mtry, was selected in an internal five-fold cross-validation of the training set with 100 randomizations. The mtry parameter represents the number of features randomly sampled from the available features at a question point in the classification tree (i.e. called splits of nodes) that, when answered, lead to the greatest improvement in classification. Six values of mtry were tested and the value that provided the largest AUROC or R^2^ was selected. We trained the random forest model using the selected mtry value and predicted the held-out test set. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance. We used AUROC and R^2^ as the prediction performance metric for classification and regression models, respectively. We used the randomForest R package (version `r installed.packages()["randomForest" ,"Version"]`) as implemented in the caret R package (version `r installed.packages()["caret" ,"Version"]`) for developing and testing our models.

@@

**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Sze_SCFACRC_mBio_2019/.

@@

## Figures

@@

@@

@@
